---
title: "Reproducible Research: Peer Assessment 1"
author: "Matthew Finney"
date: "April 21, 2017"
output: 
  html_document:
    keep_md: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
**Review Criteria 1:** Code for reading in the dataset and/or processing the data  
```{r loadingdata}
activity <- read.csv('activity.csv')
```

## What is mean total number of steps taken per day?

1. First, calculate the total number of steps taken for each day.

```{r}
totalStepsByDay <- with(activity, tapply(steps,date,sum, na.rm=TRUE))
```

**Review Criteria 2:** Histogram of the total number of steps taken each day  
2. Then, produce a histogram with the total number of steps per day.
```{r}
hist(totalStepsByDay, xlab = 'Total number of steps', main = 'Histogram of total number of steps taken each day')
```

**Review Criteria 3:** Mean and median number of steps taken each day  
3. The **mean** number of steps taken by day is `r mean(totalStepsByDay)`. The **median** number of steps taken by day is `r median(totalStepsByDay)`.


## What is the average daily activity pattern?
**Review Criteria 4:** Time series plot of the average number of steps taken  
1. Make a time series plot of the 5-minute interval and number of steps taken in that interval, averaged across all days.

```{r}
meanStepsByInterval <- with(activity, tapply(steps,interval,mean, na.rm=TRUE))

plot(rownames(meanStepsByInterval),meanStepsByInterval, type='l', xlab = '5-minute interval', ylab = 'Average Steps Taken', main = 'Average daily activity pattern')
```

**Review Criteria 5:** The 5-minute interval that, on average, contains the maximum number of steps  
2. The `r names(meanStepsByInterval[which.max(meanStepsByInterval)])` interval contains the maximum number of steps on average across all the days in the dataset.


## Imputing missing values
1. The total number of missing values in the dataset is `r sum(is.na(activity))`.

2. Missing values will be filled in with the mean number of steps for the corresponding 5 minute interval across all days, per the calculation describedi in the code chunck in the following paragraph.

**Review Criteria 6:** Code to describe and show a strategy for imputing missing data  
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activity2 <- activity
activity2$steps <- ifelse(is.na(activity2$steps), meanStepsByInterval[as.character(activity2$interval)],activity2$steps)
```

**Review Criteria 7:** Histogram of the total number of steps taken each day after missing values are imputed
4. Make a histogram of the total number of steps taken each day.
```{r}
totalStepsByDay2 <- with(activity2, tapply(steps,date,sum, na.rm=TRUE))
hist(totalStepsByDay2, xlab = 'Total number of steps', main = 'Histogram of total number of steps taken each day')
```

The **mean** number of steps taken by day is `r mean(totalStepsByDay2)`. The **median** number of steps taken by day is `r median(totalStepsByDay2)`. These values are different from the values in part 1 of the assignment. Imputing missing data on the estimates of the total daily number of steps appears to have brought the dataset into closer alignment with a standard normal curve.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
activity2$weekday <- ifelse(weekdays(as.Date(activity2$date)) %in% c('Saturday', 'Sunday'), 'weekend', 'weekday')
activity2$weekday <- as.factor(activity2$weekday)
```

**Review Criteria 8:** Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends  
2. Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.
```{r}
require(ggplot2)
require(plyr)

meanStepsByInterval2 <- ddply(activity2,.(interval,weekday),summarize, meanSteps = mean(steps))

patternPlot <- ggplot(meanStepsByInterval2, aes(y=meanSteps,x=interval)) + geom_line() + facet_grid(weekday ~ .) + xlab('Interval') +ylab('Average number of steps') + ggtitle('Average number of steps for each interval, split by weekday/weekend')

print(patternPlot)
```

**Weekdays** appear to be more active than **weekends**.

**Review Criteria 9:** All of the R code needed to reproduce the results (numbers, plots, etc.) in the report  
See the above, or RepData_PeerAssessment1.R
```{r eval = FALSE}
source(RepData_PeerAssessment1.R)
```